---
title: "Seasonal migration changes"
format: 
  html:
    toc: true
    self-contained: true
editor: visual
---

### How has the distance between the spring and fall centers of distribution changed over time?

Is there a significant difference between the distance between centroids prior to and after 2010?

```{r}
#| label: load libraries and dependencies
#| echo: false
#| message: false
#| warning: false

library(here)
library(tidyverse)
library(broom)
library(gt)
library(gmRi)
library(matrixStats)

# species data
species <- read_csv("/Users/clovas/Documents/Decadal_Variability/Data/species_list_final.csv") # more extensive species list

# seasonal distance data
seasonal_dist <- read_rds(here("Data", "seasonal_dist.rds")) %>%
  filter(count == 53) %>% # only species present in all years (excluding 2020)
  select(comname, data) %>% 
  unnest(data) %>% 
  unnest(data) %>% 
  group_by(comname, est_year) %>%
  filter(n_distinct(season) == 2)  %>%
  ungroup() # Assuming there are 2 seasons in total (Spring, Fall) ungroup()

# trawl year day
delta_year_day <- read_rds(here("Data","trawl_yearday.rds"))

# linear modz (from movement.qmd)
mods <- read_rds(here("Data", "quantiles_lm.rds"))
```

```{r}
#| label: t-test
#| echo: false
#| message: false
#| warning: false

means <- seasonal_dist %>% 
  filter(!est_year == 2017) %>% 
  select(comname, est_year, dist_km) %>%
  mutate(group = ifelse(est_year <= 2009, "1970-2009", "2010-2023")) %>% 
  group_by(comname, group) %>% 
  summarise(avg_dist = mean(dist_km)) %>% 
  pivot_wider(names_from = group, values_from = avg_dist) %>% 
  arrange(desc(`1970-2009`)) # used to check t.test estimates 

t.test <- seasonal_dist %>% 
  filter(!est_year == 2017) %>% 
  select(comname, est_year, dist_km) %>%
  mutate(group = ifelse(est_year <= 2009, "1970-2009", "2010-2023")) %>% 
  group_by(comname) %>% 
  nest() %>% 
  mutate(t.test = map(data, function(x){
    t.test(dist_km ~ group, data = x) %>% 
      tidy() %>% 
      select(
        estimate1970to2009 = estimate1, 
        estimate2010to2023 = estimate2, 
        method, 
        p.value) %>% 
      mutate(different = ifelse(p.value <= 0.05, T, F))
  })) %>% 
  unnest(t.test)

migrators <- t.test %>% 
  filter(estimate1970to2009 >= 100) %>% 
  arrange(desc(estimate1970to2009))

signif <- t.test %>% 
  filter(different == TRUE) %>% 
  mutate(direction = ifelse(estimate2010to2023 - estimate1970to2009 > 0, "+", "-"))

t.test %>% 
  select(comname, estimate1970to2009, estimate2010to2023, different) %>% 
  arrange(desc(estimate1970to2009)) %>% 
  arrange(desc(different)) %>% 
  mutate(comname = str_to_sentence(comname)) %>% 
  gt(groupname_col = NULL) 

```

```{r}
#| label: ecodata
#| echo: false
#| message: false
#| warning: false

library(ecodata)

species_groupings <- ecodata::species_groupings %>% 
  mutate(comname = tolower(COMNAME))

soe_24 <- species_groupings %>% 
  select(comname, SOE.24) %>% 
  filter(comname %in% species$comname) %>% 
  distinct() 

```

## Assessing changes in seasonal distance based on movement rates of functional groups 

```{r}
#| label: rates of change
#| echo: false
#| message: false
#| warning: false

planktivores <- mods %>% 
  left_join(soe_24) %>% 
  filter(quantile == "50%" & SOE.24 == "Planktivore") %>% 
  unnest(tidy) %>% 
  filter(!term == "(Intercept)") %>% 
  select(comname, season, SOE.24, estimate) %>%
  pivot_wider(names_from = "season", values_from = "estimate") %>% 
  mutate(spring_changes = ifelse(`Spring` > `Fall`, T,F))


piscivores <- mods %>% 
  left_join(soe_24) %>% 
  filter(quantile == "50%" & SOE.24 == "Piscivore") %>% 
  unnest(tidy) %>% 
  filter(!term == "(Intercept)") %>% 
  select(comname, season, SOE.24, estimate) %>%
  pivot_wider(names_from = "season", values_from = "estimate") %>% 
  mutate(spring_changes = ifelse(`Spring` > `Fall`, T,F))

# are piscivores able to keep pace with planktivores? 
# difference of squares...? 

```


## Plotting linear mixed model



`dist_km ~ delta_year_day + est_year*SOE_24 + (1|year_fac) + (1|comname), family = Gamma(link = "log"), REML = T)`

```{r}
#| label: mixed model
#| echo: false
#| message: false
#| warning: true

library(lme4)
library(ggeffects)
library(glmmTMB)
library(emmeans)

seasonal_dist %>% 
  # filter(comname %in% signif$comname & !est_year == 2017) %>% 
  left_join(soe_24) %>% 
  filter(!SOE.24 == "Benthos") %>% 
  left_join(delta_year_day %>% select(est_year, delta_year_day)) %>% 
  mutate(year_fac = factor(est_year)) -> lme

# We narrowed down the species to those which had a statistically significant difference between their 1970-2009 seasonal distance and their 2010-2023 seasonal distance. Size of the point represents the difference between the two averages; color represents feeding guild;x axis represents magnitude and direction of the modeled changes in seasonal distance. The model is as follows

# species_model <- glmmTMB(dist_km ~ delta_year_day + est_year*comname + (1|year_fac), family = Gamma(link = "log"), data = lme, REML = T)

species_model <- glmmTMB(dist_km ~ delta_year_day + est_year*SOE.24 + (1|year_fac) + (1|comname), data = lme, family=Gamma(link="log"))

ggpredict(species_model, terms = c("est_year", "SOE.24")) %>% 
  plot()

# summary(species_model)

# species_trends <- as.data.frame(emtrends(species_model, "comname", "est_year")) %>% 
#   left_join(soe_24) %>%
#   left_join(signif %>% group_by(comname) %>% summarise(diff = estimate2010to2023-estimate1970to2009))
# 
# ggplot(species_trends) +
#   geom_vline(xintercept = 0.0, color = "#535353", alpha = 0.7) +
#   geom_point(aes(y = forcats::fct_reorder(comname, est_year.trend), x = est_year.trend, color = SOE.24, size = abs(diff))) + 
#   geom_segment(aes(x = asymp.LCL, xend = asymp.UCL, y = forcats::fct_reorder(comname, est_year.trend)), alpha = 0.5) +
#   xlim(c(-0.05, 0.05)) +
#   xlab("slope") + ylab("species") +
#   scale_color_gmri() + 
#   theme_gmri(panel.grid.major = element_line(linetype = 1, color = "#e9e9e9"),
#              panel.grid.minor = element_line(linetype = 1, color = "#e9e9e9"),
#              panel.border = element_rect(linetype = 1, color = "black")) 

```
